Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2013-01-22 HRUIonGPU/jrnl_article.tex
uffc article

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging. For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures. For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer. Since the execution time highly depends on the number of elements in the given array, it is important to find ways of reducing computational complexity for high-channel-count systems. Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we have selected to implement the beamspace transformation, described by Nilsen and Hafizovic \cite{Nilsen2009}, on the GPU to maintain high framerates as the channel count increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
\end{align}
The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Even if different windows can be used in different areas of the images they have traditionally been predefined.

@@

The Capon beamformer can be summerized in three main steps. First a sample covariance matrix has to be calculated, 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} (\vec{x}_l^{\Delta}[n'])(\vec{x}_l^{\Delta}[n'])^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray of length $L$ $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$ and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L}
\end{align}
has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

Second, a linear system of equations has to be solved
\begin{align}\label{eq:b}
\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
\end{align}
where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed.

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $\text{CONV}_m$ is a non-truncated convolution across elements.
Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited. The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}. In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.

@@

An SM has available a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. It is also important that the problem can be divided into enough threads to hide instruction latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.

@@

\section{Parallel Capon}\label{sec:meth}
If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively. As $L$ increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
&- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H). \label{eq:sliding}
\end{align}
The same is true in time where
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?). However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}). In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$. Note that sliding will break independence along the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$. We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU. This also means that we do not rely on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.

Diagonal loading as a constant value can be added with marginal cost. For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When it is found, scattered writes of the diagonal elements to global memory has to be performed.

@@

For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$. The solver is open for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$. Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r}^{\Delta} = \sum_{k=r}^{r+N_L-1}x_{k}^{\Delta} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}^{\Delta}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore easily remove beams where no interference is present by removing rows from $\mat{B}$.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray is given as:
\begin{align}\label{eq:bs_formula}
x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$. If adaptive diagonal loading is applied after data has been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams. Thus the average energy per beam is higher than the average energy per element, if the system uses focused transmit beams. This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.

@@

Fig. \ref{fig:phantom} presents a phantom simulated using Field II \cite{Jensen1992}. The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array. The transducer was exited with a $3\lambda$ long pulse and the transmit focus is at 80 mm. The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB. Speckle level is at 0 dB. The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. This simulation is presented in an attached video.

Fig. \ref{fig:invivo} shows an \textit{in-vivo} image of the left ventricle. The image was acquired using a experimental scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode. The image is extracted from a cardiac cycle, that can be viewed in its entirety in the attached videos.

Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$. These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images. For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.

>>>>

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging. For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures. For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer. Since the execution time highly depends on the number of elements in the given array, it is important to find ways of reducing computational complexity for high-channel-count systems. Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of the beamspace transformation, described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the channel count increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam in space. This delayed data is denoted $x_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR). And even if different windows can be used in different areas of the images they have traditionally been predefined.

@@

The Capon beamformer can be summerized in three main steps. First a sample covariance matrix has to be calculated, 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$ and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L}
\end{align}
has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

Second, a linear system of equations has to be solved
\begin{align}\label{eq:b}
\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
\end{align}
where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $\text{CONV}_m$ is a non-truncated convolution across elements.
Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited. Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}. In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.

@@

An SM has available a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. It is also important that the problem can be divided into enough threads to hide instruction latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.

@@

\section{Parallel Capon}\label{sec:meth}
If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively. As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
The same is true in time where
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?). However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$. Note that sliding will break independence along the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the GPU compute threads per $\mat{\hat{R}}$ are organized in such a way that the lower triangle can be calculated with little overhead. This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.

Diagonal loading as a constant value can be added with negligible cost. For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When it is found, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$. The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$. Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore easily remove beams where no interference is present by removing rows from $\mat{B}$.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
\begin{align}\label{eq:bs_formula}
x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$. If adaptive diagonal loading is applied after data have been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams. Thus the average energy per beam is higher than the average energy was per element, if the system uses focused transmit beams. This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.

@@

Fig. \ref{fig:phantom} presents a phantom simulated using Field II \cite{Jensen1992}. The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array. The transducer was exited with a $3\lambda$ long pulse and the transmit focus is at 80 mm. The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB. Speckle level is at 0 dB. The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in an attached video.

Fig. \ref{fig:invivo} shows an \textit{in-vivo} image of the left ventricle. The image was acquired using a experimental scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode. The image is extracted from a cardiac cycle, that can be viewed in its entirety in an attached video.

Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$. These settings have previously been shown to provide high resolution while maintaining DAS-like speckle in ultrasound images \cite{Synnevag2007a}. For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.
